Supp mat: time series paper

### 2018-02-21 writing/manuscript_draft_short_version.Rmd
changes to short format version of the article

---
bibliography: White_bib.bib
fontsize: 12pt
geometry: margin=1in
header-includes: \usepackage{float} \usepackage{lineno} \usepackage{setspace}\doublespacing \usepackage[firstinits=true,style=authoryear-comp,natbib=true,doi=false,isbn=false,url=false,uniquename=false,uniquelist=false,sorting=none]{biblatex} 
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
  word_document: default
editor_options: 
  chunk_output_type: console
---
Running head: Power to detect population trends [33 of 40 characters]

@@

To be submitted to: \emph{Ecology} as a Statistical Report
\vspace{1 mm}

Number of pages: 20  
\vspace{1 mm}

@@

Number of references: 33
\vspace{1 mm}

@@

Data and code for all the figures and tables can be found at (https://github.com/erwhite1/time-series-project).
\vspace{2 mm}

@@

\begin{abstract}
Long-term time series are necessary to better understand population dynamics, assess species' conservation status, and make management decisions. However, population data are often expensive, requiring a lot of time and resources. When is a population time series long enough to address a question of interest? I determine the minimum time series length required to detect significant increases or decreases in population abundance. To address this question, I examine 822 populations of vertebrate species. I show that on average 15.9 years of continuous monitoring are required in order to achieve a high level of statistical power. However, there is a wide distribution around this average, casting doubt on simple rules of thumb. The minimum time required depends on trend strength, population variability, and temporal autocorrelation. However, there were no life-history traits (e.g. generation length) that were predictive of the minimum time required. These results point to the importance of sampling populations over long periods of time. I argue that statistical power needs to be considered in monitoring program design and evaluation. Short time series are likely under-powered and potentially misleading.
\end{abstract}

Keywords: ecological time series, experimental design, monitoring, power analysis, statistical power, sampling design

@@

Observational studies and population time series have become a cornerstone of modern ecological research and conservation biology [@Magurran2010; @Hughes2017]. Long-term data are necessary to both understand population dynamics and to assess species extinction risk. Some time series may now be considered "long-term" (e.g. continuous plankton recorder, @Giron-Nava2017), but most are still short. Time series are typically short due to short funding cycles and typical experimental time-frames [@Field2007;@Hughes2017].

@@

Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Wagner2009; @Giron-Nava2017]. For example, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates. Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error. @Rueda-Cediel2015 also used a modeling approach, but parameterized a model specific for a threatened snail, \emph{Tasmaphena lamproides}. They found that for this short-lived organism, 15 years was adequate to assess long-term trends in abundance. However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on only a few species.

I estimate the minimum time series length required ($T_{min}$) to assess long-term changes in abundance via simple linear regression. Statistical power is not a new tool [@Cohen1992], but it is still under-appreciated in ecological research [@Legg2006]. Therefore, I only briefly describe the idea of statistical power for time series analyses using simulations. I then examine 822 population time series to estimate $T_{min}$.

@@

# Methods

## Statistical power in time series

An important step in experimental design is to determine the number of samples required. For any particular experiment four quantities are intricately linked: significance level ($\alpha$), statistical power, effect size, and sample size [@Legg2006]. The exact relationship between these quantities depends on the specific statistical test. A type I error is a false positive, or incorrect rejection of a true null hypothesis. For example, if a time series was assessed as significantly increasing or decreasing---when there was no true significant trend---this would be a false positive. The false positive rate, or significance level ($\alpha$) is often set at 0.05 (although this is purely historical, @Mapstone1995). A type II error ($\beta$) is a failure to detect a true trend, or failure to reject a false null hypothesis. Formally, statistical power ($1-\beta$) is one minus the probability of a type II error ($\beta$). The effect size is a estimate of the strength of a particular phenomenon. Prior to an experiment, one could set appropriate levels of power, significance level, and the effect size to estimate the sample size required for the experiment. This approach, however, is not straight-forward for a time series, or more complicated scenarios [@Johnson2015], as data are clearly non-independent.

## Simulation approach

One approach to determining the minimum time series length needed is through repetitive simulations of a population model [@Gerrodette1987]. This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses [@Bolker2008;@Johnson2015]. Essentially, a population model, with a selected set of parameter values, is simulated repetitively for a number of years. As an example, we can take the following population model for population size $N$ at time $t$:
\begin{equation}
N(t + 1) = N(t) + r(t) + \epsilon \mbox{ with } \epsilon \sim N(\mu, \sigma)
\label{sim_model}
\end{equation}
where $\epsilon$ is a normally-distributed random noise term with mean $\mu$ and standard deviation $\sigma$. The rate of growth ($r$) is the trend strength of the increase or decrease (i.e. the estimated slope from linear regression). It is important to note that any population model could be used here. As an example, in the supplementary material (Fig. \ref{fig:shark_example}), an age-structured model of population dynamics is used in place of equation \ref{sim_model} [@White2014].

Statistical power is the proportion of simulations where the slope parameter from linear regression is significant at the 0.05 threshold. Statistical power of 0.8 would indicate that, if there was a true trend in abundance, there would be a 0.8 probability of detecting the trend. Values of 0.05 for the significance level and 0.8 statistical power are historical [@Cohen1992] and it is important to examine the effect of changing these values (Fig. \ref{fig:min_time_vs_alpha_beta}).

In Fig. \ref{fig:theoretical_approach}a, the model in equation \ref{sim_model} is simulated repetitively. Statistical power increases with increases in the length of time sampled (Fig. \ref{fig:theoretical_approach}b). Where power is greater than 0.8 (the dotted line), that is the minimum time required ($T_{min}$) to be confident in the detection of a long-term trend in abundance. As shown previously [@Rhodes2011; @Rueda-Cediel2015], statistical power increases with larger trend strength and lower population variability (Fig. \ref{fig:theoretical_approach}c,d). Simulation approaches can be useful before designing a monitoring program or when a realistic model exists for the population in question. However, it is also informative to examine how statistical power changes with sampling effort in empirical time series.

@@

## Data source
I use a database of `r length(table(dat$ID))` population time series compiled by @Keith2015. The data are originally from the Global Population Dynamics Database [@GPDD2010] and several other sources [@Keith2015]. I filtered out short time series (less than `r pop_info$series_length[1]` years), and those with missing data, leaving `r nrow(pop_info)` time series. The data includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish. <!-- The data also includes information on generation length and survey specifications. For each time series, I also calculate other variables of interest: coefficient of variance in population size, long-term trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation.-->

@@

## Empirical approach

I assume that each time series is long enough to include all necessary information (e.g. variability) about the population. In other words, each time series is a representative sample. I first take all possible contiguous subsamples of each time series. For example, a time series of 35 years would have 34 possible contiguous subsamples of length 2, 34 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 35 [@Gerber1999; @Brashares2005; @Giron-Nava2017]. In line with the simulation approach, I determine the proportion of subsamples of a particular length that have estimated slope coefficients statistically different from zero. This proportion is a measure of statistical power. Lastly, I determine which subsample length is required to achieve a certain threshold of statistical power (0.8, @Cohen1992). The minimum subsampled length that met these criteria is the minimum time series length required ($T_{min}$). All analyses were conducted in R [@RCoreTeam2016].

@@

# Results

@@

I examined a database of `r nrow(pop_info)` separate population time series representing `r length(table(pop_info$Binomial))` species. This database consisted of vertebrate species with a variety of life-history characteristics. Across all the populations, I found an average minimum time series length required ($T_{min}$) of `r round(mean(pop_info$min_time_for_power),1)` (SD=`r round((var(pop_info$min_time_for_power))^0.5,1)`), with a wide distribution (Fig. \ref{fig:min_time_dist}b). Estimates of $T_{min}$ varied between biological class (Fig. \ref{fig:min_time_dist}a). Ray-finned fish (class Actinopterygii) typically had estimates of $T_{min}$ over 20 years. Birds (class Aves) had a much wider distribution of $T_{min}$, but usually required less years of sampling. Differences between these classes can be explained by differences in variability in population size and strength of trends in abundance (Fig. \ref{fig:class}).

@@

@@

@@

# Discussion

@@

I examined `r nrow(pop_info)` population time series. I then subsampled each to determine the minimum time required to achieve a desired significance level and power for linear regression. Statistical power is important as it provides information on the necessary number of samples required to determine a significant trend [@Legg2006]. I found that on average `r round(mean(pop_info$min_time_for_power),2)` years of continuous monitoring were typically necessary (Fig. \ref{fig:min_time_dist}b). However, the distribution of minimum time required was wide. Therefore, simplistic rules of thumb are probably not appropriate. This time-frame is in line with past work on a short-lived snail species [@Rueda-Cediel2015] and a long-lived whale species [@Gerber1999]. @Hatch2003 used seabird monitoring data to estimate minimum sampling requirements. He found that the time required ranged from 11 to 69 years depending on species, trend strength, and study design.

@@

An important related idea is the optimal allocation of sampling effort in space versus time. In a theoretical investigation of this question, @Rhodes2011 found that the optimal allocation of sampling depended strongly on temporal and spatial autocorrelation. If spatial population dynamics were highly correlated, then it was better to sample more temporally, and vice versa. My work supports this idea as populations with strong temporal autocorrelation needed less years of sampling (Fig. \ref{fig:correlates}). @Morrison2008 also studied the optimal allocation of sampling effort in space versus time, but used emprical data from a long-term survey of the collared pika (\emph{Ochotona collaris}) in the Yukon. They found that surveys less than 5 years may be misleading and that extrapolating from one population to another, even when nearby geographically, may be untenable.

@@

As a test of the consequences of not conducting power analyses, I compared the results here to the IUCN Redlist criteria. IUCN criteria A2 suggests that species that have experienced 30% declines over 10 years (or three generations) should be listed as vulnerable [@IUCN2012]. However, for the populations I examined, this criteria may be too simplistic (Fig. \ref{fig:IUCN_analysis}). For many populations, the IUCN criteria suggest more years than necessary are required to assess a population as vulnerable (points below diagonal line in Fig. \ref{fig:IUCN_analysis}). Conversely, for other populations the IUCN criteria suggest sampling times that are less than the minimum time required for statistical power. This suggests that the IUCN criteria are probably too simplistic as the minimum time required does not correlate with generation time or other biological covariates (Fig. \ref{fig:correlates}d-h).

The design of monitoring programs should include calculations of statistical power, the allocation of sampling in space versus time [@Rhodes2011], and metrics to increase precision [@Seavy2007]. Ideally, a formal decision analysis to evaluate these different factors would be conducted to design or assess any monitoring program [@Hauser2006; @McDonald-Madden2010]. This type of formal decision analysis would also include information on the costs of monitoring. These costs include the actual costs of sampling [@Brashares2005] and the ecological costs on inaction [@Thompson2000].

@@

## Limitations

This paper has some limitations in determining the minimum time series length required. First, $T_{min}$ is particular to the specific question of interest. An additional complication is that for the empirical approach, the subsampling of the full time series allows for estimates of power, but the individual subsamples are clearly not independent of one another. In an ideal setting, a specific population model would be parameterized for each population of interest. Then, model simulations could be used to estimate the minimum time series required to address each specific question of interest. Clearly, this is not always practical, especially if conducting analyses for a wide array of species. <!--In addition, the statistical models suggest that $T_{min}$ does not correlate with any life-history traits, at least for the question of linear regression (Fig. \ref{fig:correlates}d-h). Therefore, it is not possible to use these results to predict $T_{min}$ for another population, even if the population is of a species with a similar life-history to one in the database used.-->

## Conclusions

Power analyses are not a novel aspect of ecological research [@Legg2006]. However, power analyses are still underutilized, especially in the context of time series analyses. I used a database of `r nrow(pop_info)` populations to determine the minimum time series length required to detect population trends. This goes beyond previous work that either focused on theoretical investigations or a limited number of species. I show that to identify long-term changes in abundance, on average `r round(mean(pop_info$min_time_for_power),2)` years of continuous monitoring are often required (Fig. \ref{fig:min_time_dist}). However, there is wide distribution of estimated minimum times. Therefore, it is probably not wise to use a simple threshold number of years in monitoring design. Further, contrary to my initial hypotheses, minimum time required did not correlate with generation time or any other life-history traits (Fig. \ref{fig:correlates}d-h). This result argues against overly simplified measures of minimum sampling time based on generation length or other life-history traits (Fig. \ref{fig:IUCN_analysis}).

My work implies that for many populations, short time series are probably not reliable for detecting population trends. This result highlights the importance of long-term monitoring programs. From both a scientific and management perspective, estimates of $T_{min}$ are important. If a time series is too short, we lack the statistical power to reliably detect long-term population trends. In addition, a time series that is too long may be a poor use of already limited funds [@Gerber1999]. Further, more data is not always best in situations where management actions need to be taken [@Martin2012; @Martin2017]. When a population trend is detected, it may be too late for management action. In these situations, the precautionary principle may be more appropriate [@Thompson2000]. Future work should examine other species, with a wider range of life-history characteristics. In addition, similar approaches can be used to determine the minimum time series length required to address additional questions of interest.

@@

  #load command
  # Plotting commands
  par(mfrow=c(2,2),oma=c(1.5,2,0.5,0.5),mar=c(3,5,2.5,0.5))
    matplot(simulations,type='l',col=rgb(0.5,0.5,0.5,0.5),lty=1,ylab='',xlab='',las=1,cex.axis=1.2)
      mtext(text = 'time' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'population size' ,side = 2,cex = 1.2,line=3.5)
      mtext(text = '(a)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)

    plot(years_to_sample,power,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2)
      abline(v=min_time_needed,lty=2,lwd=2,col='black')
      abline(h=0.8,lty=2,lwd=2,col='black')
      mtext(text = 'time sampled' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'statistical power' ,side = 2,cex = 1.2,line=3)
      mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)

@@

      plot(parameter_combinations$r,parameter_combinations$min_time_required,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2,ylim=c(0,50))
      mtext(text = 'trend strength (slope coefficient)' ,side = 1,cex = 1.2,line=3)
      mtext(text = expression(T[min]) ,side = 2,cex = 1.2,line=3)
      mtext(text = '(c)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)

@@

      #save(parameter_combinations,file='../analysis-outputs/min_time_vs_pop_var.Rdata')
      load('../analysis-outputs/min_time_vs_pop_var.Rdata')
      plot(parameter_combinations$sigma,parameter_combinations$min_time_required,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2,ylim=c(0,50))
      mtext(text = 'population variability' ,side = 1,cex = 1.2,line=3)
      mtext(text = expression(T[min]) ,side = 2,cex = 1.2,line=3)
      mtext(text = '(d)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
```

@@

require(ggplot2)
plot1<- ggplot(aes(class,min_time_for_power),data=pop_info) + geom_violin() + geom_jitter(col=pop_info$color,height = 0, width = 0.1,pch=pop_info$shape) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black"),axis.text = element_text(size=12),axis.title.x =element_text(size=16) ,axis.title.y =element_text(size=16) ) +
 labs(x='biological class',y='minimum time required') +
  geom_text(aes(x=4.4,y=33),label='(a)',size=6) +
  scale_y_continuous(limits = c(0, 40))

@@

  plot(pop_info$gen_len,pop_info$min_time_for_power,pch=pop_info$shape,col=pop_info$color,las=1,ylab='',xlab='',cex.axis=1.2,cex=1.2)
  abline(lm(LPI_pop_info$min_time_for_power~LPI_pop_info$gen_len),lwd=3)
  mtext(text = 'generation length (years)',side = 1,line = 3,cex = 1.2)
  mtext('D',3,line=0,adj = 0.0,cex=1.2)

>>>>

---
bibliography: White_bib.bib
csl: ecology-letters.csl
fontsize: 12pt
geometry: margin=1in
header-includes: 
  \usepackage{float} 
  \usepackage{lineno} 
  \usepackage{setspace}\doublespacing 
  \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{,}
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: no
  word_document: default
editor_options: 
  chunk_output_type: console
---
<!--Running head: Power to detect population trends-->

@@

To be submitted to: \emph{BioScience} as a Forum
\vspace{1 mm}

Number of words: 4,400 (21 pages)  
\vspace{1 mm}

@@

Number of references: 34
\vspace{1 mm}

@@

All data and code is freely available at: (https://github.com/erwhite1/time-series-project).
\vspace{2 mm}

@@

\begin{abstract}
Long-term time series are necessary to better understand population dynamics, assess species' conservation status, and make management decisions. However, population data are often expensive, requiring a lot of time and resources. What is the minimum population time series length required to detect significant trends in abundance? I first present an overview of the theory and past work that has tried to address this question. As a test of these approaches, I then examine 822 populations of vertebrate species. I show that on average 15.9 years of continuous monitoring are required in order to achieve a high level of statistical power. However, there is a wide distribution around this average, casting doubt on commonly used, simple rules of thumb. These results point to the importance of long-term monitoring. I argue that statistical power needs to be considered more often in monitoring programs. Short time series are likely under-powered and potentially misleading.
\end{abstract}

Keywords: ecological time series, experimental design, population monitoring, statistical power, sampling design

@@

Observational studies and population time series have become a cornerstone of modern ecological research and conservation biology [@Peters2010; @Magurran2010; @Hughes2017]. Long-term data are necessary to both understand population dynamics and to assess species extinction risk. Some time series may now be considered "long-term" (e.g. continuous plankton recorder, @Giron-Nava2017), but most are still short. Time series are typically short due to short funding cycles and typical experimental time-frames [@Field2007;@Hughes2017].

@@

Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Wagner2009; @Giron-Nava2017]. For example, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates. Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error. @Rueda-Cediel2015 also used a modeling approach, but parameterized a model specific for a threatened snail, \emph{Tasmaphena lamproides}. They found that for this short-lived organism, 15 years was adequate to assess long-term trends in abundance. However, these studies, and other past work, have typically been only on theoretical aspects of monitoring design or focused on a few species.

Statistical power is not a new tool [@Gibbs1998; @Cohen1992], but it is still under-appreciated in ecological research [@Legg2006]. Therefore, I begin by reviewing key concepts of power analyses in relation to time series analysis. I then explain how simulation approaches have been used to estimate the minimum time required to estimate long-term population trends. Lastly, I take an empirical approach to estimate the minimum time required for 822 animal populations.

@@

# Statistical power in time series analyses

An important step in experimental design is to determine the number of samples required. For any particular experiment four quantities are intricately linked: significance level ($\alpha$), statistical power, effect size, and sample size [@Legg2006]. The exact relationship between these quantities depends on the specific statistical test. A type I error is a false positive, or incorrect rejection of a true null hypothesis. For example, if a time series was assessed as significantly increasing or decreasing---when there was no true significant trend---this would be a false positive. The false positive rate, or significance level ($\alpha$) is often set at 0.05 (although this is purely historical, @Mapstone1995). A type II error ($\beta$) is a failure to detect a true trend, or failure to reject a false null hypothesis. Formally, statistical power ($1-\beta$) is one minus the probability of a type II error ($\beta$). The effect size is a estimate of the strength of a particular phenomenon.

Prior to an experiment, one could set appropriate levels of power, significance level, and the effect size to estimate the sample size required for the experiment. This approach, however, is not straight-forward for a time series, or more complicated scenarios [@Johnson2015], as data are clearly non-independent.

In the context of time series data, sample size can be the number of study sites surveyed, frequency of surveys per year, and the number of years surveyed. For example, @Gibbs1998 examined how many times within a year a population needs to be sampled to ensure high statistical power. They found that the sampling intensity within a year differed greatly depending on species, because of differences in population variability. I explore similar methods, but I focus on the number of years required to estimate trends in abundance.

<!--For time series data, simulations can be designed for a specific population and question [@Bolker2008; @Johnson2015]. Simple models can be simulated with parameter values corresponding to a population of interest [@Gerrodette1987]. Statistical power is the proportion of simulations that then meet some set of criteria. The specific criteria depend on the question at hand. For example, given a time series, when is the slope from linear regression significantly different from zero? In other words, when is the time series significantly increasing or decreasing? It is then possible to determine how power changes with a variable of interest. For example, time series can be simulated for different lengths of time. From these simulations, the minimum time series length required to meet certain levels of statistical significance and power is estimated [@Bolker2008]. -->

# Simulation approach

One approach to determining the minimum time series length needed is through repetitive simulations of a population model [@Gerrodette1987;@Gibbs1998]. This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses [@Bolker2008;@Johnson2015]. Essentially, a population model, with a selected set of parameter values, is simulated repetitively for a number of years. As an example, we can take the following population model for population size $N$ at time $t$:
\begin{equation}
N(t + 1) = N(t) + r(t) + \epsilon \mbox{ with } \epsilon \sim N(\mu, \sigma)
\label{sim_model}
\end{equation}
where $\epsilon$ is a normally-distributed random noise term with mean $\mu$ and standard deviation $\sigma$. The rate of growth ($r$) is the trend strength of the increase or decrease (i.e. the estimated slope from linear regression).

With regard to detecting time series trends, statistical power is the proportion of simulations where the slope parameter from linear regression is significant at the 0.05 threshold. Statistical power of 0.8 would indicate that, if there was a true trend in abundance, there would be a 0.8 probability of detecting the trend. Values of 0.05 for the significance level and 0.8 statistical power are historical [@Cohen1992] and it is important to examine the effect of changing these values (Fig. \ref{fig:min_time_vs_alpha_beta}).

In Fig. \ref{fig:theoretical_approach}a, the model in equation \ref{sim_model} was simulated repetitively. Statistical power increases with increases in the length of time sampled (Fig. \ref{fig:theoretical_approach}b). Where power is greater than 0.8 (the dotted line), that is the minimum time required ($T_{min}$) to be confident in the detection of a long-term trend in abundance. As shown previously [@Rhodes2011; @Rueda-Cediel2015], statistical power increases with larger trend strength and lower population variability (Fig. \ref{fig:theoretical_approach}c,d). Simulation approaches can be useful before designing a monitoring program or when a realistic model exists for the population in question. However, it is also informative to examine how statistical power changes with sampling effort in empirical time series.

It is important to note that any population model could be used here (see example in Fig. \ref{fig:min_time_dist_log_pop}). Ideally, the specific model choice should be tailored to the population of interest. As an example, I determined the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014]. I found that over 27 years of continuous monitoring were needed in this particular scenario (Fig. \ref{fig:shark_example}). Not surprisingly, the minimum time required for the lemon shark population was strongly dependent on model parameters.  Similarly, @Rueda-Cediel2015 used a matrix model parameterized for a particular snail species. They used the model to argue that only 10-15 years were needed to accurately assess trends in abundance.

@@

# Empirical approach

As an empirical test of these ideas, I used a database of `r length(table(dat$ID))` population time series compiled by @Keith2015. The data are originally from the Global Population Dynamics Database [@GPDD2010] and several other sources [@Keith2015]. I filtered out short time series (less than `r pop_info$series_length[1]` years), and those with missing data, leaving `r nrow(pop_info)` time series. The data includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish. <!-- The data also includes information on generation length and survey specifications. For each time series, I also calculate other variables of interest: coefficient of variance in population size, long-term trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation.-->

@@

I assumed that each time series was long enough to include all necessary information (e.g. variability) about the population. In other words, each time series was a representative sample. I first took all possible contiguous subsamples of each time series. For example, a time series of 35 years had 34 possible contiguous subsamples of length 2, 34 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 35 [@Gerber1999; @Brashares2005; @Giron-Nava2017]. In line with the simulation approach, I determined the proportion of subsamples of a particular length that had estimated slope coefficients statistically different from zero. This proportion is a measure of statistical power. Lastly, I determined which subsample length is required to achieve a certain threshold of statistical power (0.8, @Cohen1992). The minimum subsampled length that met these criteria was the minimum time series length required ($T_{min}$). All analyses were conducted in R [@RCoreTeam2017].

@@

<!--RESULTS-->

@@

## Estimates of the minimum time required

<!--I examined a database of `r nrow(pop_info)` separate population time series representing `r length(table(pop_info$Binomial))` species. This database consisted of vertebrate species with a variety of life-history characteristics.--> Across all the populations, I found an average minimum time series length required ($T_{min}$) of `r round(mean(pop_info$min_time_for_power),1)` (SD=`r round((var(pop_info$min_time_for_power))^0.5,1)`), with a wide distribution (Fig. \ref{fig:min_time_dist}b). Estimates of $T_{min}$ varied between biological class (Fig. \ref{fig:min_time_dist}a). Ray-finned fish (class Actinopterygii) typically had estimates of $T_{min}$ over 20 years. Birds (class Aves) had a much wider distribution of $T_{min}$, but usually required less years of sampling. Differences between these classes can be explained by differences in variability in population size and strength of trends in abundance (Fig. \ref{fig:class}).

This time-frame is in line with past work on a short-lived snail species [@Rueda-Cediel2015] and a long-lived whale species [@Gerber1999]. @Hatch2003 used seabird monitoring data to estimate minimum sampling requirements. He found that the time required ranged from 11 to 69 years depending on species, trend strength, and study design.

@@

I initially hypothesized that species with longer lifespans or generation times may require a longer sampling period. This result could have been a result of at least two factors. First, the data I used may not include a diverse enough set of species with different life-history traits. Second, the question I posed, whether a population is increasing or decreasing, was specifically concerned with population trends over time. Therefore, life-history characteristics may be more important for other questions more closely tied to species biology. For example, @Blanchard2007 used detailed simulations of spatially-distributed fisheries to compare survey designs. They found that statistical power depended on survey design, temperature preferences, and the degree of population patchiness.

@@

For the populations I examined, the IUCN criteria may be overly simplistic (Fig. \ref{fig:IUCN_analysis}). For many populations, the IUCN criteria suggest more years than necessary are required to assess a population as vulnerable. Conversely, for other populations the IUCN criteria suggest sampling times that are less than the minimum time required for statistical power. This suggests that the IUCN criteria are probably too simplistic as the minimum time required does not correlate with generation time or other biological covariates (Fig. \ref{fig:correlates}d-h).

@@

# Related questions

@@

<!--
I examined `r nrow(pop_info)` population time series. I then subsampled each to determine the minimum time required to achieve a desired significance level and power for linear regression. Statistical power is important as it provides information on the necessary number of samples required to determine a significant trend [@Legg2006]. I found that on average `r round(mean(pop_info$min_time_for_power),2)` years of continuous monitoring were typically necessary (Fig. \ref{fig:min_time_dist}b). However, the distribution of minimum time required was wide. Therefore, simplistic rules of thumb are probably not appropriate. This time-frame is in line with past work on a short-lived snail species [@Rueda-Cediel2015] and a long-lived whale species [@Gerber1999]. @Hatch2003 used seabird monitoring data to estimate minimum sampling requirements. He found that the time required ranged from 11 to 69 years depending on species, trend strength, and study design.

@@

An important related idea is the optimal allocation of sampling effort in space versus time. In a theoretical investigation of this question, @Rhodes2011 found that the optimal allocation of sampling depended strongly on temporal and spatial autocorrelation. If spatial population dynamics were highly correlated, then it was better to sample more temporally, and vice versa. The empirical data supports this idea as populations with strong temporal autocorrelation needed less years of sampling (Fig. \ref{fig:correlates}). @Morrison2008 also studied the optimal allocation of sampling effort in space versus time, but used emprical data from a long-term survey of the collared pika (\emph{Ochotona collaris}) in the Yukon. They found that surveys less than 5 years may be misleading and that extrapolating from one population to another, even when nearby geographically, may be untenable.

@@

<!--
As a test of the consequences of not conducting power analyses, I compared the results here to the IUCN Redlist criteria. IUCN criteria A2 suggests that species that have experienced 30% declines over 10 years (or three generations) should be listed as vulnerable [@IUCN2012]. However, for the populations I examined, this criteria may be too simplistic (Fig. \ref{fig:IUCN_analysis}). For many populations, the IUCN criteria suggest more years than necessary are required to assess a population as vulnerable (points below diagonal line in Fig. \ref{fig:IUCN_analysis}). Conversely, for other populations the IUCN criteria suggest sampling times that are less than the minimum time required for statistical power. This suggests that the IUCN criteria are probably too simplistic as the minimum time required does not correlate with generation time or other biological covariates (Fig. \ref{fig:correlates}d-h). 
-->

@@

# Limitations

This paper has some limitations in determining the minimum time series length required. First, the minimum time estimated is particular to the specific question of interest. An additional complication is that for the empirical approach, the subsampling of the full time series allows for estimates of power, but the individual subsamples are clearly not independent of one another. In an ideal setting, a specific population model would be parameterized for each population of interest. Then, model simulations could be used to estimate the minimum time series required to address each specific question of interest. Clearly, this is not always practical, especially if conducting analyses for a wide array of species. <!--In addition, the statistical models suggest that $T_{min}$ does not correlate with any life-history traits, at least for the question of linear regression (Fig. \ref{fig:correlates}d-h). Therefore, it is not possible to use these results to predict $T_{min}$ for another population, even if the population is of a species with a similar life-history to one in the database used.-->

# Conclusions

Power analyses are not a novel aspect of ecological research [@Legg2006]. However, power analyses are still underutilized, especially in the context of time series analyses. I used a database of `r nrow(pop_info)` populations to determine the minimum time series length required to detect population trends. This goes beyond previous work that either focused on theoretical investigations or a limited number of species. I show that to identify long-term changes in abundance, on average `r round(mean(pop_info$min_time_for_power),2)` years of continuous monitoring are often required (Fig. \ref{fig:min_time_dist}). However, there is wide distribution of estimated minimum times. Therefore, it is probably not wise to use a simple threshold number of years in monitoring design. Further, contrary to my initial hypotheses, minimum time required did not correlate with generation time or any other life-history traits (Fig. \ref{fig:correlates}d-h). This result argues against overly simplified measures of minimum sampling time based on generation length or other life-history traits, like those of the IUCN criteria (Fig. \ref{fig:IUCN_analysis}).

The design of monitoring programs should include calculations of statistical power, the allocation of sampling in space versus time [@Larsen2001; @Rhodes2011], and metrics to increase precision [@Seavy2007]. Ideally, a formal decision analysis to evaluate these different factors would be conducted to design or assess any monitoring program [@Hauser2006; @McDonald-Madden2010]. This type of formal decision analysis would also include information on the costs of monitoring. These costs include the actual costs of sampling [@Brashares2005] and the ecological costs of inaction [@Thompson2000].

For many populations, short time series are probably not reliable for detecting population trends. This result highlights the importance of long-term monitoring programs. From both a scientific and management perspective, estimates of $T_{min}$ are important. If a time series is too short, we lack the statistical power to reliably detect long-term population trends. In addition, a time series that is too long may be a poor use of already limited funds [@Gerber1999]. Further, more data is not always best in situations where management actions need to be taken [@Martin2012; @Martin2017]. When a population trend is detected, it may be too late for management action. In these situations, the precautionary principle may be more appropriate [@Thompson2000]. Future work should examine other species, with a wider range of life-history characteristics. In addition, similar approaches can be used to determine the minimum time series length required to address additional questions of interest.

@@

  #load command
  # Plotting commands
  par(mfrow=c(2,2),oma=c(1.5,2,0.5,0.5),mar=c(3,5,2.5,0.5))
    matplot(simulations,type='l',col=rgb(0.5,0.5,0.5,0.5),lty=1,ylab='',xlab='',las=1,cex.axis=1.2)
      mtext(text = 'Time' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'Population size' ,side = 2,cex = 1.2,line=3.5)
      mtext(text = '(a)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)

    plot(years_to_sample,power,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2)
      abline(v=min_time_needed,lty=2,lwd=2,col='black')
      abline(h=0.8,lty=2,lwd=2,col='black')
      mtext(text = 'Time sampled' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'Statistical power' ,side = 2,cex = 1.2,line=3)
      mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)

@@

      plot(parameter_combinations$r,parameter_combinations$min_time_required,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2,ylim=c(0,50))
      mtext(text = 'Trend strength (slope coefficient)' ,side = 1,cex = 1.2,line=3)
      mtext(text = expression(T[min]) ,side = 2,cex = 1.2,line=3)
      mtext(text = '(c)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)

@@

      #save(parameter_combinations,file='../analysis-outputs/min_time_vs_pop_var.Rdata')
      load('../analysis-outputs/min_time_vs_pop_var.Rdata')
      plot(parameter_combinations$sigma,parameter_combinations$min_time_required,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2,ylim=c(0,50))
      mtext(text = 'Population variability' ,side = 1,cex = 1.2,line=3)
      mtext(text = expression(T[min]) ,side = 2,cex = 1.2,line=3)
      mtext(text = '(d)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)
```

@@

require(ggplot2)
plot1<- ggplot(aes(class,min_time_for_power),data=pop_info) + geom_violin() + geom_jitter(col=pop_info$color,height = 0, width = 0.1,pch=pop_info$shape) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),panel.background = element_blank(), axis.line = element_line(colour = "black"),axis.text = element_text(size=12),axis.title.x =element_text(size=16) ,axis.title.y =element_text(size=16) ) +
 labs(x='Biological class',y='Minimum time required') +
  geom_text(aes(x=4.4,y=33),label='(a)',size=6) +
  scale_y_continuous(limits = c(0, 40))

@@

  plot(pop_info$gen_len,pop_info$min_time_for_power,pch=pop_info$shape,col=pop_info$color,las=1,ylab='',xlab='',cex.axis=1.2,cex=1.2)
  abline(lm(LPI_pop_info$min_time_for_power~LPI_pop_info$gen_len),lwd=3)
  mtext(text = 'Generation length (years)',side = 1,line = 3,cex = 1.2)
  mtext('D',3,line=0,adj = 0.0,cex=1.2)
